---
title: "Introduction to hullBase"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hullBase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
```

```{r setup}
library(hullBase)
```

## Overview

hullBase removes the baseline from spectra such as FTIR spectra using the
rubberband method. The rubberband baseline is the lower convex hull of
the spectrum, which works well for flat, sloped and bowl shaped
baselines. It cannot follow a concave (dome shaped) baseline, because a
convex hull can only bend upwards. Over a dome it draws a straight line
instead, and part of the background is left in the corrected spectrum.

This package adds a fix for that limitation. It finds the parts of the
hull that hide a dome and corrects only those parts by local bending.

## Example data

`simulate_spectrum()` builds a test spectrum with a known baseline,
following the simulation design of Zhang et al. (2020). The spectrum is
the sum of Gaussian peaks, a baseline and Gaussian noise.

$$y = \sum_k a_k \exp\left[-\left(\frac{x - \mu_k}{w_k}\right)^2\right] + b(x) + \varepsilon$$

Here the baseline is a concave arch and the noise level is set by a
signal to noise ratio of 30 dB.

```{r data}
set.seed(1)
sim <- simulate_spectrum(baseline = "concave", snr_db = 30)
head(sim)
```

## Correcting a spectrum

`correct()` runs the full method and returns a `BaselineModel` object.
The only setting is `peak_width`, the width of the widest peak or group
of overlapping peaks.

```{r correct}
fit <- correct(sim, peak_width = 300)
fit
summary(fit)
```

The plot shows the plain rubberband baseline (red, dashed) and the
baseline from the adaptive method (blue). The shaded part is where the
correction was applied.

```{r plot}
plot(fit)
plot(fit, which = "corrected")
```

Because the true baseline is known, the error of each method can be
measured with the root mean square error (RMSE).

```{r rmse}
rmse <- function(b) sqrt(mean((b - sim$baseline)^2))
c(rubberband = rmse(fit$rubberband), adaptive = rmse(fit$baseline))
```

## How the method works

**1. Noise level.** `noise_estimate()` uses the median absolute
deviation of the second differences. A second difference has noise
variance $6\sigma^2$, so

$$\hat\sigma = \frac{\mathrm{MAD}(y_{i+1} - 2y_i + y_{i-1})}{\sqrt 6}$$

**2. Rubberband baseline.** `lower_hull()` finds the lower convex hull
with the monotone chain algorithm (Andrew 1979) and joins the hull points
with straight lines.

**3. Finding concave segments.** For each straight segment of the hull,
`detect_concave_segments()` takes the gap $g_i$ between the data and the
segment and applies a moving minimum over a window as wide as
`peak_width`.

$$e_i = \min_{|j - i| \le m} g_j, \qquad m = \mathrm{round}\left(\frac{\texttt{peak\_width}}{2h}\right)$$

Peaks are narrower than the window, so they are removed. A dome is wider,
so it remains. A segment is flagged when the largest $e_i$ is above
$2\hat\sigma\sqrt{2\ln n}$, twice the universal threshold of Donoho and
Johnstone (1994).

**4. Local bending.** `refine_segment()` adds a bowl $q(x)$ that is zero
at both ends of the segment, takes the lower hull, and removes the bowl
again. This is the bending idea of Beleites (2015), applied automatically
to one segment at a time.

$$q(x) = \frac{(x - x_L)(x - x_R)}{(W/2)^2}, \qquad b = \mathrm{hull}(y + cq) - cq$$

The bend strength $c$ is twice the dome height estimated from $e_i$. A
parabola needs a strength equal to its height, and other dome shapes need
more, so the factor of 2 gives a margin.

## Choosing peak_width

`peak_width` should cover the widest peak or group of overlapping peaks.
For a Gaussian peak this is about 2.5 times its full width at half
maximum. If it is too small, overlapping peaks are mistaken for a dome.

```{r peak-width}
set.seed(3)
convex <- simulate_spectrum(baseline = "convex")
sum(detect_concave_segments(convex$x, convex$y, peak_width = 100)$flagged)
sum(detect_concave_segments(convex$x, convex$y, peak_width = 300)$flagged)
```

With `peak_width = 100` two segments are wrongly flagged. With 300 none
are.

## Results

Average RMSE over 10 simulated spectra for each baseline shape and noise
level.

```{r results}
set.seed(2026)
results <- data.frame()
for (shape in c("concave", "mixed", "flat", "linear", "convex")) {
  for (snr in c(30, 25)) {
    err <- replicate(10, {
      s <- simulate_spectrum(baseline = shape, snr_db = snr)
      f <- correct(s, peak_width = 300)
      c(sqrt(mean((f$rubberband - s$baseline)^2)),
        sqrt(mean((f$baseline - s$baseline)^2)))
    })
    results <- rbind(results, data.frame(shape = shape, snr_db = snr,
                                         rubberband = round(mean(err[1, ]), 3),
                                         adaptive = round(mean(err[2, ]), 3)))
  }
}
results
```

For concave and mixed baselines the error is much smaller. For flat,
linear and convex baselines nothing is flagged, so the result is the same
as the plain rubberband.

## Compiled code

The hull and the moving minimum are loops where each step depends on the
previous one, so they are written in C++ with Rcpp. The moving minimum
uses a monotone queue (Lemire 2006), so its speed does not depend on the
window size. Plain R versions are kept in the package for comparison.

```{r speed}
x <- seq(400, 4000, length.out = 7200)
y <- sin(x / 500) + rnorm(7200, 0, 0.01)
time_it <- function(f) system.time(for (i in 1:50) f())[["elapsed"]]
c(hull_R = time_it(function() hullBase:::lower_hull_r(x, y)),
  hull_Cpp = time_it(function() hullBase:::lower_hull_cpp(x, y)),
  erosion_R = time_it(function() hullBase:::erode(y, 150)),
  erosion_Cpp = time_it(function() hullBase:::erode_cpp(y, 150)))
```

Times are in seconds for 50 runs.

## Limitations

* A peak wider than `peak_width` looks the same as a dome and is partly
  removed as background.
* The hull rests on the lowest noise values, so the baseline sits about
  two noise standard deviations too low. This is true for any rubberband
  baseline.
* Inside a bent segment, a wide peak can be reported a few percent too
  low.
* The spectrum should start and end on baseline, because the first and
  last points always lie on the hull.

## References

Andrew, A. M. (1979). Another efficient algorithm for convex hulls in two
dimensions. Information Processing Letters, 9(5), 216 to 219.

Beleites, C. (2015). Fitting baselines to spectra. hyperSpec package
vignette.

Donoho, D. L. and Johnstone, I. M. (1994). Ideal spatial adaptation by
wavelet shrinkage. Biometrika, 81(3), 425 to 455.

Lemire, D. (2006). Streaming maximum minimum filter using no more than
three comparisons per element. Nordic Journal of Computing, 13(4),
328 to 339.

Zhang, F., Tang, X., Tong, A., Wang, B. and Wang, J. (2020). An automatic
baseline correction method based on the penalized least squares method.
Sensors, 20, 2015.
